---
title: "PNAS_Delivery"
author: "BJC"
date: "May 15, 2015"
output: html_document
---

Initialize:
```{r init, message=FALSE}
library("phyloseq"); packageVersion("phyloseq")
library("ggplot2"); packageVersion("ggplot2")
library("gridExtra")
library("igraph"); packageVersion("igraph")
set.seed(100)
default.par <- par(no.readonly = TRUE)
options(stringsAsFactors = FALSE)
theme_set(theme_bw())
```

Load data:
```{r load-data, message=FALSE}
###This supposes all files have been downloaded to
###a pregnancy folder just under the user's main directory
setwd("~/Pregnancy")
otu_file <- "PregnancyClosed15.RData"
load(otu_file)
```

## Alpha-diversity Disruption after Delivery

Let's look at alpha diveristy vs. time relative to delivery, with samples separated by whether they are collected before or after delivery status. Also perform a paired Wilcoxon test for a significant change in diversity after delivery:
```{r fig5}
g_lmes <- list()
g_lms <- list()
s_lms <- list()
s_lmes <- list()
disp <- list(Vaginal_Swab="Vagina", Saliva="Saliva", Stool="Stool", Tooth_Gum="Tooth/Gum")
for(site in sites) {
  # Estimate diversity and make data frame
  ps <- PSbs[[site]]
  ps <- prune_samples(sample_data(ps)$GWColl >= 10, ps)
  ps <- prune_samples(sample_data(ps)$GWColl <= 80, ps)
  pptab <- table(sample_data(ps)$SubjectID, sample_data(ps)$PostPreg)>0
  keep <- (pptab[,1] & pptab[,2])
  keep <- names(keep[keep])
  ps <- prune_samples(sample_data(ps)$SubjectID %in% keep, ps)
  sample_data(ps)$WeeksRelDel <- sample_data(ps)$GWColl - sample_data(ps)$GWDel
  erDF <- estimate_richness(ps, split = TRUE, measures = c("Shannon"))
  df <- data.frame(erDF, sample_data(ps))
  
  # LM modeling/plotting
  p <- ggplot(df, aes(x = WeeksRelDel, y = Shannon, color=PostPreg))
  p <- p + geom_point(size=0.5, color="black", na.rm = TRUE)
  #p <- p + theme(axis.text.x = element_text(angle = -90, vjust = 0.5, 
  #                                         hjust = 0))
  p = p + ylab("Shannon index")
  #p = p + facet_wrap(~variable, nrow = nrow, scales = scales)
  p <- p + xlab("Time relative to delivery (weeks)")
  p <- p + ylim(0, 4.5)
  p <- p + guides(color=FALSE)
  p <- p + geom_vline(x_intercept=0, linetype="dashed")
  p <- p + theme(axis.title.x = element_text(size=8),
                axis.title.y = element_text(size=8),
                axis.text.x = element_text(size=7),
                axis.text.y = element_text(size=7),
                plot.title = element_text(size=8),
                plot.margin=unit(c(0,1.7,0,1.7),"mm"))
  g_lms[[site]] <- p + stat_smooth(method="lm") + ggtitle(disp[[site]])
  
  # Keep the samples immediately before and after delivery
  subdfs <- split(df, df$SubjectID)
  subs <- names(subdfs)
  
  for(sub in subs) {
    subdf <- subdfs[[sub]]
    if((sum(subdf$GWColl > subdf$GWDel) == 0) || (sum(subdf$GWColl < subdf$GWDel) == 0)) {
      subdfs[[sub]] <- NULL
    } else {
      # Keep sample before and after delivery
      before <- which(subdf$GDColl == max(subdf$GDColl[subdf$GDColl < subdf$GDDel]))
      after <- which(subdf$GDColl == min(subdf$GDColl[subdf$GDColl > subdf$GDDel]))
      subdfs[[sub]] <- subdf[c(before, after),]
    }
  }
  # Combine the subdfs back together
  dfba <- do.call(rbind,subdfs)
  Utest <- wilcox.test(dfba[!dfba$PostPreg,"Shannon"], dfba[dfba$PostPreg, "Shannon"], paired=T)
  print(site)
  print(Utest)
}
library("gridExtra")
do.call(grid.arrange,c(g_lms, nrow=2))  # FIGURE 5
Fig5 <- do.call(arrangeGrob, c(g_lms, nrow=2))  # FIGURE 5
##To save a pdf version.
#ggsave(Fig5, file = "PregPNAS_Fig5.pdf", width=8.45, height=7, units=c("cm"))
```

Look at difference for vaginal community between Csection/Traditional delivery:
```{r see-Csect}
for(site in sites) {
  ps <- PSbs[[site]]
  ps <- prune_samples(sample_data(ps)$GWColl >= 10, ps)
  ps <- prune_samples(sample_data(ps)$GWColl <= 80, ps)
  pptab <- table(sample_data(ps)$SubjectID, sample_data(ps)$PostPreg)>0
  keep <- (pptab[,1] & pptab[,2])
  keep <- names(keep[keep])
  ps <- prune_samples(sample_data(ps)$SubjectID %in% keep, ps)
  
  sample_data(ps)$WeeksRelDel <- sample_data(ps)$GWColl - sample_data(ps)$GWDel
  sample_data(ps)$DeliveryType <- c("Vaginal")
  sample_data(ps)[sample_data(ps)$CSection,"DeliveryType"] <- c("C-Section")
  
  erDF <- estimate_richness(ps, split = TRUE, measures = c("Shannon"))
  df <- data.frame(erDF, sample_data(ps))
  
  # LM modeling/plotting
  p <- ggplot(df, aes(x = WeeksRelDel, y = Shannon, color=PostPreg))
  p <- p + geom_point(color="black", size=1, na.rm = TRUE)
  #p <- p + theme(axis.text.x = element_text(angle = -90, vjust = 0.5, 
  #                                         hjust = 0))
  p = p + ylab("Alpha-diversity (Shannon)")
  #p = p + facet_wrap(~variable, nrow = nrow, scales = scales)
  p <- p + xlab("Time relative to delivery (weeks)")
  p <- p + ylim(0, 4.5)
  p <- p + guides(color=FALSE)
  p <- p + geom_vline(x_intercept=0, linetype="dashed")
  p <- p + stat_smooth(method="lm")
  print(p + facet_wrap(~DeliveryType))
  print(paste(site, "Subject numbers"))
  print(apply(table(sample_data(ps)$SubjectID, sample_data(ps)$DeliveryType) > 0, 2, sum))
  #+ theme(title=element_text(size=rel(1.5)))
}
```

```{r}
ps <- PSbs[["Vaginal_Swab"]]
ps <- prune_samples(sample_data(ps)$PostPreg, ps)
foo <- table(sample_data(ps)$SubjectNice)
foo
ps <- prune_samples(sample_data(ps)$SubjectNice %in% names(foo)[foo>=3], ps)
print(ps)

erDF <- estimate_richness(ps, split = TRUE, measures = c("Shannon"))
df <- data.frame(erDF, sample_data(ps))
# Annoying re-ordering because @sample_data doesn't save ordered factors
np <- 7; nm <- 5; nt <- 28
df$SubjectNice <- ordered(df$SubjectNice, levels = paste0(c(rep("P", np), rep("M",nm), rep("T",nt)), c(seq(1,np), seq(1,nm), seq(1,nt))))
df$WeeksPastDel <- df$GWColl - df$GWDel

erDFpreg <- estimate_richness(PSPreg[["Vaginal_Swab"]], split = TRUE, measures = c("Shannon"))
samdf <- data.frame(sample_data(PSPreg[["Vaginal_Swab"]]))
pregdivs <- tapply(erDFpreg$Shannon, samdf$SubjectNice, mean)
df$PregMeanShannon <- sapply(df$SubjectNice, function(x) pregdivs[[x]])
#df$SubjectID <- ordered(df$SubjectID, levels = as.character(newnames))
p <- ggplot(df, aes(x=WeeksPastDel, y=Shannon))
p <- p + geom_line() + facet_wrap(~SubjectNice)
p <- p + geom_hline(aes(yintercept=PregMeanShannon), linetype="dashed")
p + xlab("Time past delivery (weeks)") + ylab("Alpha-diversity (Shannon)")
```
