---
title: "Stability and Homogeneity"
author: "BJC"
date: "April 9, 2015"
output: html_document
---

Stability and Homogeneity
========================================================
This Rmd file tests for a progressive change in the microbial communities at each body site over the course of pregnancy. Three properties of these communities were tested: Alpha-diversity, Instability (as characterized by the distance between consecutive samples within a subject), and Beta-diversity (between subjects). Figure 1 and the related Supplementary Figures 2-5 are herein produced, and all the statistical tests related to progressive change during pregnancy are herein performed.

Initialize:
```{r init, message=FALSE}
library("phyloseq"); packageVersion("phyloseq")
library("ggplot2"); packageVersion("ggplot2")
#library("igraph"); packageVersion("igraph")
library("nlme"); packageVersion("nlme")
library("reshape2"); packageVersion("nlme")
library("gridExtra")
set.seed(100)
default.par <- par(no.readonly = TRUE)
options(stringsAsFactors = FALSE)
theme_set(theme_bw())
ANSIZE=2   # Plot visual parameter
PVALLEN=5  # Plot visual parameter
```

Load data:
```{r load-data, message=FALSE}
setwd("~/Pregnancy")
otu_file <- "PregnancyClosed15.RData"
load(otu_file)
```

# Alpha-diversity During Pregnancy

## Calculate alpha-diversity

Calculate the diversities, using three metrics (Shannon, Chao1, Simpson), for each sample during pregnancy, keeping only those >=10 weeks (since almost no subjects provided samples before then):
```{r}
meths = c("Shannon", "Chao1", "Simpson")
divdfs <- list()

for(site in sites) {
  # Estimate diversity and make data frame
  ps <- PSPreg[[site]]
  ps <- prune_samples(sample_data(ps)$GWColl >= 10, ps)
  erDF <- estimate_richness(ps, split = TRUE, measures = meths)
  df <- data.frame(erDF, sample_data(ps))
  for(meth in meths) { df[,paste0("log", meth)] <- log(df[,meth]) }
  divdfs[[site]] <- df
}
```

Histogram the distributions of the calculated diversity metrics:
```{r warning=FALSE}
df <- do.call(rbind, divdfs)
df <- melt(df, id.vars=c("SampleID","BodySite"), measure.vars=c("Chao1", "Shannon", "Simpson"))
p <- ggplot(data=df, aes(x=value))
p <- p + geom_histogram(aes(y=..count..))
p + facet_grid(BodySite ~ variable, scales="free")
```

## Normality and transforms

For Shannon and Chao1, normality looks acceptable except for the Vaginal Swab site. Simpson is more questionable, but since Shannon is our main focus we'll let that be. The crucial thing is that the Vaginal Swab diversity strongly deviates from normality, therefore lets try to transform it to improve normality before fitting. Checking the key part:

```{r log-vag-hist}
dfv <- df[df$BodySite=="Vaginal_Swab",]
dfv$log_value <- log(dfv$value)
dfv$sqrt_value <- sqrt(dfv$value)
pnone <- ggplot(data=dfv, aes(x=value))
pnone + geom_histogram(aes(y=..count..)) + facet_wrap(~variable, scales="free")
plog <- ggplot(data=dfv, aes(x=log_value))
plog + geom_histogram(aes(y=..count..)) + facet_grid(~variable, scales="free")
psqrt <- ggplot(data=dfv, aes(x=sqrt_value))
psqrt + geom_histogram(aes(y=..count..)) + facet_grid(~variable, scales="free")
```

While not perfect, the log-transform clearly improved the normality substantially. Given the previous use of log-transformed shannon vaginal diversity in the literature, we'll go ahead and use this for futher analysis of this data:

## LME modeling

We use a linear-mixed-effects model for the relationship of alpha-diversity and gestational time, with a Subject-dependent intercept/slope as the random effect:  
```{r model-alpha}
# Plotting parameters for each measure
pvs <- list(Shannon=list(ymin=0, ymax=5, ylab="Alpha-diversity (Shannon)", xan = 30, yan=4.8),
           Chao1=list(ymin=0, ymax=300, ylab="Alpha-diversity (Chao1)", xan = 30, yan=288),
           Simpson=list(ymin=0, ymax=1.2, ylab="Alpha-diversity (Simpson)", xan = 30, yan=1.15))

ggs_alpha <- list() # List of ggplot graphs for each site/method
models_alpha <- list() # List of returns from the lme(...) model fitting function

meths = c("Shannon", "Chao1", "Simpson")
for(site in sites) {
  df <- divdfs[[site]]
  for(meth in meths) {
    pv <- pvs[[meth]]
    key <- paste(site, meth)
    
    # MLE modeling/plotting
    new.dat <- data.frame(GWColl=10:40)
    mlef <- formula(paste(meth, "~", "GWColl"))
    if(site == "Vaginal_Swab") mlef <- formula(paste0("log", meth, "~ GWColl"))
    model.mx <- lme(mlef, random = ~ 1+GWColl|SubjectID, data=df)
    new.dat$pred <- predict(model.mx, newdata=new.dat,level=0)
    
    Designmat <- model.matrix(eval(eval(model.mx$call$fixed)[-2]), new.dat[-ncol(new.dat)])
    
    #compute standard error for predictions
    predvar <- diag(Designmat %*% model.mx$varFix %*% t(Designmat))
    new.dat$SE <- sqrt(predvar)
    new.dat$hi2SE <- new.dat$pred + 2*new.dat$SE
    new.dat$lo2SE <- new.dat$pred - 2*new.dat$SE

    # Reverse the log-transform for the vaginal data
    if(site == "Vaginal_Swab") {
      new.dat$hi2SE <- exp(new.dat$pred + 2*new.dat$SE)
      new.dat$lo2SE <- exp(new.dat$pred - 2*new.dat$SE)
      new.dat$pred <- exp(new.dat$pred)
    }
    
    #Extract p-value
    pval <- summary(model.mx)$tTable["GWColl", "p-value"]
    
    # Build ggplot with confidence interval
    p1 <- ggplot(new.dat,aes(x=GWColl,y=pred)) + 
      geom_line(color="blue") +
      geom_ribbon(aes(ymin=lo2SE,ymax=hi2SE),alpha=0.2,fill="blue") +
      geom_point(data=df,aes_string(y=meth), size=0.5) +
      xlab("Gestational Time (Weeks)") +
      ggtitle(site) + 
      ylim(pv$ymin, pv$ymax)
    
    # Special plot options for Tooth/Gum
    if(site=="Tooth_Gum" & meth=="Chao1") { # More digits on this plot
      p1 <- p1 + annotate("text", x=pv$xan, y=pv$yan, size=ANSIZE, label = paste("p =", substr(as.character(pval),1,6)))
    } else {
      p1 <- p1 + annotate("text", x=pv$xan, y=pv$yan, size=ANSIZE, label = paste("p =", substr(as.character(pval),1,PVALLEN)))
    }

    # Save ggplot in list
    p1 <- p1 + ylab(NULL)
    ggs_alpha[[key]] <- p1

    # Summarize the LME model fit and save in list
#    print(paste0(site, ": LME {nlme}"))
#    print(summary(model.mx))
    models_alpha[[key]] <- model.mx
  }
}

# Show plot of the 3x4 models (3 methods, 4 sites) performed above
listorder <- paste(sites, rep(meths, each=4))
ggs_alpha <- ggs_alpha[listorder]
do.call(grid.arrange,c(ggs_alpha, ncol=4, main=paste("Diversity vs. Gestational Time\nLME Modeling --", "Shannon,", "Chao1,", "Simpson")))
```

There is no significant fixed effect for a trend (slope) of diversity during pregnancy at any of the body sites.

# Stability During Pregnancy

As a proxy for community stability (or instability) we use the distance between consecutive samples of a community within a subject. As we are interested in trends during pregnancy, we will perform an analogous analysis on this measure of instability as we did on alpha-diversity above.

## Calculate instability

First we define a couple functions to extract the distances between consecutive samples:
```{r}
# Function to pull out SampleIDs by the GWColl
getSamByGW <- function(x, cc) {
  if(x %in% cc$GWColl) { 
    return(cc$SampleID[which(x==cc$GWColl)][[1]])
  } else { return(NA) }
}

# Function to calculate distance between two samples
getDist2Sam <- function(sam1, sam2, ps, ...) {
  phyloseq::distance(prune_samples(c(sam1, sam2), ps), ...)[[1]]
}

# Reduce down to those with previous samples
samdat.prune_prev <- function(samdat) {
  samdf <- data.frame(samdat)
  subjects <- unique(samdf$SubjectID)
  csub <- split(samdf, samdf$SubjectID)
  for(sub in subjects) {
    cc <- csub[[sub]]
    cc$Prev <- ((cc$GWColl-1) %in% cc$GWColl)
    cc$PrevID <- sapply(cc$GWColl-1, function(x) getSamByGW(x, cc))
    if(sum(cc$Prev) == 0) {
      csub[[sub]] <- NULL
    } else {
      cc <- cc[cc$Prev,]
      csub[[sub]] <- cc
    }
  }
  return(do.call(rbind, csub))
}
```

Calculate distances between consecutive time points:
```{r calc-prev-dists, warning=FALSE}
###This requires more memory and multiple CPUS
###managed by packages doParallel and foreach
library("doParallel")
library("foreach")
registerDoParallel(makeCluster(3))

samdf_prevs <- list()
for(site in sites) {
  ps <- PSPreg[[site]]
  ps <- prune_samples(sample_data(ps)$GWColl >= 10, ps)
  ps <- transform_sample_counts(ps, function(OTU) OTU/sum(OTU))
  samdf_prevs[[site]] <- samdat.prune_prev(sample_data(ps))
  rownames(samdf_prevs[[site]]) <- samdf_prevs[[site]]$SampleID
  samdf_prev <- samdf_prevs[[site]]
  samdf_prevs[[site]]$PrevBray <- mapply(getDist2Sam, samdf_prev$SampleID,
        samdf_prev$PrevID, MoreArgs = list(ps=ps, method="bray"))
  samdf_prevs[[site]]$logPrevBray <- log(samdf_prevs[[site]]$PrevBray)
  samdf_prevs[[site]]$PrevJSD <- mapply(getDist2Sam, samdf_prev$SampleID,
        samdf_prev$PrevID, MoreArgs = list(ps=ps, method="jsd"))
  samdf_prevs[[site]]$logPrevJSD <- log(samdf_prevs[[site]]$PrevJSD)
  samdf_prevs[[site]]$PrevWUF <- mapply(getDist2Sam, samdf_prev$SampleID,
        samdf_prev$PrevID, MoreArgs = list(ps=ps, method="unifrac", weighted=T, parallel=T))
  samdf_prevs[[site]]$logPrevWUF <- log(samdf_prevs[[site]]$PrevWUF)
}
```


## Normality and transforms

Histogram the distributions of the instability measure:
```{r see-hists-2}
df <- do.call(rbind, samdf_prevs)
df <- melt(df, id.vars=c("SampleID","BodySite"), measure.vars=c("PrevBray", "PrevJSD", "PrevWUF"))
p <- ggplot(data=df, aes(x=value))
p <- p + geom_histogram(aes(y=..count..))
p + facet_grid(variable~BodySite, scales="free")
```

Normality looks acceptable except for the Vaginal Swab site, again. We'll again try to transform it to improve normality before fitting. Checking the key part:
```{r log-vag-hist-2}
dfv <- df[df$BodySite=="Vaginal_Swab",]
dfv$log_value <- log(dfv$value)
dfv$sqrt_value <- sqrt(dfv$value)
p <- ggplot(data=dfv, aes(x=value))
p + geom_histogram(aes(y=..count..)) + facet_wrap(~variable, scales="free")
p2 <- ggplot(data=dfv, aes(x=log_value))
p2 + geom_histogram(aes(y=..count..)) + facet_wrap(~variable, scales="free")
p3 <- ggplot(data=dfv, aes(x=sqrt_value))
p3 + geom_histogram(aes(y=..count..)) + facet_wrap(~variable, scales="free")
```

Once again log is obvioulsy a major improvement in normality, if not perfect.

## MLE modeling

Fit and plot the distances between subsequent time points as a function of gestational weeks (GWColl):
```{r plot-prev-dist}
ggs_instability <- list()  # List of ggplots of instability vs. gestational time
models_instability <- list() # List of return values from the lme(...) function
meths <- c("Bray", "JSD", "WUF") # Distances to use
# pvs: A list of graphical parameters for plotting
pvs <- list(Bray=list(ymin=0, ymax=1, ylab="Instability (Bray)", xan = 30, yan=1),
           JSD=list(ymin=0, ymax=1, ylab="Instability (JSD)", xan = 30, yan=1),
           WUF=list(ymin=0, ymax=0.7, ylab="Instability (WUF)", xan = 30, yan=0.67))
sites <- c("Vaginal_Swab", "Saliva", "Stool", "Tooth_Gum")

for(site in sites) {
  df <- samdf_prevs[[site]]
  if(site == "Tooth_Gum") df <- df[-which(df$SampleID == "1002304148"),] 
  # A mystery, lme hangs unless this specific sample is removed...
  # "nlminb problem, convergence error code = 1
  #  message = singular convergence (7)"
  for(meth in meths) {
    pv <- pvs[[meth]]
    key <- paste(site, meth)

    # MLE modeling/plotting
    lmeControl(maxIter=500, msMaxIter=500)
    lmef <- formula(paste(paste0("Prev",meth), "~", "GWColl"))
    if(site == "Vaginal_Swab") lmef <- formula(paste0("logPrev", meth, "~ GWColl"))
    model.mx <- lme(lmef, random = ~ 1+GWColl|SubjectID, data=df,
                    control=lmeControl(maxIter=500, msMaxIter=500))
    new.dat <- data.frame(GWColl=10:40)
    new.dat$pred <- predict(model.mx, newdata=new.dat,level=0)
    
    #create design matrix
    Designmat <- model.matrix(eval(eval(model.mx$call$fixed)[-2]), new.dat[-ncol(new.dat)])
    
    #compute standard error for predictions
    predvar <- diag(Designmat %*% model.mx$varFix %*% t(Designmat))
    new.dat$SE <- sqrt(predvar)
    new.dat$hi2SE <- new.dat$pred + 2*new.dat$SE
    new.dat$lo2SE <- new.dat$pred - 2*new.dat$SE
#    new.dat$SE2 <- sqrt(predvar+model.mx$sigma^2)

    # Reverse log-transform for the vaginal community
    if(site == "Vaginal_Swab") {
      new.dat$hi2SE <- exp(new.dat$pred + 2*new.dat$SE)
      new.dat$lo2SE <- exp(new.dat$pred - 2*new.dat$SE)
      new.dat$pred <- exp(new.dat$pred)
    }
    
    #Extract p-value
    pval <- summary(model.mx)$tTable["GWColl", "p-value"]

    # Build ggplot with confidence interval 
    p1 <- ggplot(new.dat,aes(x=GWColl,y=pred)) + 
      geom_line(color="red") +
      geom_ribbon(aes(ymin=lo2SE,ymax=hi2SE),alpha=0.2,fill="red") +
      geom_point(data=df,aes_string(y=paste0("Prev",meth)), size=0.5) +
      xlab("Gestational Time (Weeks)") +
      annotate("text", x=pv$xan, y=pv$yan, size=ANSIZE, label = paste("p =", substr(as.character(pval),1,PVALLEN))) +
      ylim(pv$ymin, pv$ymax)
    
    # Save ggplot in a list
    p1 <- p1 + ylab(NULL)
    ggs_instability[[key]] <- p1
    
    # Summarize the LME model and save to a list
#    print(paste0(site, ": LME {nlme}"))
#    print(summary(model.mx))
    models_instability[[key]] <- model.mx
  }
}
listorder <- paste(sites, rep(meths, each=length(sites)))
ggs_instability <- ggs_instability[listorder]
do.call(grid.arrange, c(ggs_instability, ncol=4, main=paste("Stability vs. Gestational Time\nLME Modeling --", "Bray,", "JSD,", "WUF")))
```

There is no significant fixed effect for a trend (slope) of instability during pregnancy at any of the body sites.


# Beta-diversity During Pregnancy

We now consider if there is any change in beta-diversity over the course of prengnacy, where beta-diversity is defined here as the average distance between subjects (at that gestational time).

## Calculate beta-diversity and test for a significant trend

Define functions:
```{r}
require("doParallel")
require("foreach")
registerDoParallel(makeCluster(3))
# Function to pull out SampleIDs by the GWColl
getSamByGW <- function(x, cc) {
  if(x %in% cc$GWColl) { 
    return(cc$SampleID[which(x==cc$GWColl)][[1]])
  } else { return(NA) }
}

# Function to calculate distance between two samples
getDist2Sam <- function(sam1, sam2, ps, ...) {
  phyloseq::distance(prune_samples(c(sam1, sam2), ps), ...)[[1]]
}

GWSTART=10;GWEND=40
gws <- seq(GWSTART, GWEND)

# Function to get lm coeffs for the beta-diversity vs. gw fit
get_mb <- function(ps, gwdists=NULL, permute=F) {
  samdf <- data.frame(sample_data(ps))
  subdfs <- split(samdf, samdf$SubjectID)
  
  if(permute) {
    for(sub in names(subdfs)) {
      if(runif(1) > 0.5) {
        # Reversing in the GW10-40 range
        subdfs[[sub]]$GWColl <- (GWSTART+GWEND-subdfs[[sub]]$GWColl)
      }
    }
  }
  
  betas <- seq_along(gws)
  names(betas) <- gws
  for(gw in gws) {
    sams <- sapply(subdfs, function(x) getSamByGW(gw, x))
    sams <- sams[!is.na(sams)]
    if(is.null(gwdists)) {
      if(meth=="weighted unifrac") {
        dists <- phyloseq::distance(prune_samples(sams, ps), method="unifrac", weighted=T, parallel=T)
      } else {
        dists <- phyloseq::distance(prune_samples(sams, ps), method=meth, parallel=T)
      }
    } else {
      dists <- gwdists[[as.character(gw)]]
      dists <- as.matrix(dists)
#      print(gw)
#      print(sams)
      dists <- dists[sams, sams]
      dists <- as.dist(dists)
    }
    
    betas[[as.character(gw)]] <- mean(dists)
  }
  lmo <- lm(beta ~ gw, data.frame(beta=betas, gw=gws))
  m <- lmo$coefficients[["gw"]]
  b <- lmo$coefficients[["(Intercept)"]]
  return(list(m=m, b=b, lmo=lmo))
}
```

Calculate beta-diversities and estimate the trend between beta-diversity and gestational time. Show this estimated trend against the null distribution of trends estimated from the ensemble of randomly reversed time-courses. Determine pvalue by comparison to that null distribution (takes perhaps tens of minutes to run):
```{r}
meths <- c("bray", "jsd", "wuf")
par(mfrow=c(2,2))
NPERM=1000
betadivs <- data.frame()
mbs <- data.frame()
mbs_perm <- data.frame()
for(meth in meths) {
  for(site in sites) {
    ps <- PSPreg[[site]]
    ps <- transform_sample_counts(ps, function(OTU) OTU/sum(OTU))
    samdf <- data.frame(sample_data(ps))
    
    # Calculate the distance matrix for each pair of complementary gws
    getDist <- function(gw, meth) phyloseq::distance(prune_samples(samdf$GWColl %in% c(gw, GWSTART+GWEND-gw), ps), method=meth, parallel=T)
    getWUF <- function(gw) phyloseq::distance(prune_samples(samdf$GWColl %in% c(gw, GWSTART+GWEND-gw), ps), method="unifrac", weighted=T, parallel=T)
    if(meth == "wuf") gwdists <- lapply(gws, getWUF)
    else gwdists <- lapply(gws, getDist, meth=meth)
    names(gwdists) <- gws
    
    fit <- get_mb(ps, gwdists=gwdists, permute=F)
    m <- fit$m;b <- fit$b
    
    # Do the time-reversal permutations
    pms <- rep(-999, NPERM); pbs <- rep(-999, NPERM)
    for(ii in seq(1,NPERM)) {
      pfit <- get_mb(ps, gwdists=gwdists, permute=T)
      pms[[ii]] <- pfit$m
      pbs[[ii]] <- pfit$b
    }
    
    # plot the slope (m) estimated from the data along with the histograms of ms from
    # the permutations. Show pval calculated from that.
    betas <- fit$lmo$model$beta
    plot(gws, betas, ylab=meth, main=site)
    abline(a=fit$b, b=fit$m, col="blue")
    pval <- mean(pms < -abs(m)) + mean(pms > abs(m))
    hist(pms, breaks=30, main=paste(site, meth, "p:", pval))
    abline(v=m, col="red", lw=3)
    
    # Save the linear fit params and calculated beta-diversities
    mbs <- rbind(mbs, data.frame(site=site, meth=meth, m=m, b=b, pval=pval))
    mbs_perm <- rbind(mbs_perm, data.frame(site=site, meth=meth, m=pms, b=pbs))
    betadivs <- rbind(betadivs, data.frame(site=site, meth=meth, GWColl=gws, beta=betas))
  }
}
par(default.par)
```

Maybe there is a trend towards more beta-diversity in the stool samples? Let's make better plots:

```{r plot-beta-hists, warning=FALSE}
hbeta <- ggplot(data=mbs_perm, aes(x=m)) + geom_histogram()
hbeta <- hbeta + geom_vline(data=mbs, aes(xintercept=m), color="red", linetype="dashed")
hbeta <- hbeta + geom_text(x=0.0058, y=0.3*NPERM, size=3, aes(label=pval), data=mbs)
hbeta <- hbeta + facet_grid(meth~site)
hbeta
```

No. No significant trends at any of the body sites for any of the three distance metrics.

## Plot beta-diversity trend over time with uncertainty band

Use the ensemble slopes/intercepts estimated from the random time-reversals to calculate the "error" at each time-point in the linear fit of the beta-diversity (similar to "stat_smooth"):
```{r plot-beta-trends}
meths <- c("bray", "jsd", "wuf")
preds <- t(mapply(function(m,b) m*gws+b, mbs_perm$m, mbs_perm$b))
colnames(preds) <- as.character(gws)
df_perm <- cbind(mbs_perm, preds)
dfs <- split(df_perm, list(df_perm$site, df_perm$meth))
betadivs$lo <- c(NA)
betadivs$hi <- c(NA)
split_betas <- split(betadivs, list(betadivs$site, betadivs$meth))
split_mbs <- split(mbs, list(mbs$site, mbs$meth))
for(site in sites) {
  for(meth in meths) {
    key <- paste0(site,".",meth)
    df <- dfs[[key]]
    predvar <- apply(df[,as.character(gws)], 2, var)
    lo <- apply(df[,as.character(gws)], 2, function(x) quantile(x, 0.025))
    hi <- apply(df[,as.character(gws)], 2, function(x) quantile(x, 0.975))
    pred <- split_mbs[[key]][[1,"m"]]*gws + split_mbs[[key]][[1,"b"]]
    split_betas[[key]]$lo <- lo
    split_betas[[key]]$hi <- hi
    split_betas[[key]]$pred <- pred
    split_betas[[key]]$SE <- sqrt(predvar)
  }
}
beta_divs <- unsplit(split_betas, list(betadivs$site, betadivs$meth))
```

Use the calculated errors to plot the linear fit with an associated confidence band:
```{r}
meths <- c("bray", "jsd", "wuf")
pvs <- list(bray=list(ymin=0, ymax=1, ylab="Beta-diversity (Bray)", xan = 30, yan=1),
           jsd=list(ymin=0, ymax=1, ylab="Beta-diversity (JSD)", xan = 30, yan=1),
           wuf=list(ymin=0, ymax=0.5, ylab="Beta-diversity (WUF)", xan = 30, yan=0.48))

b_fits <- list()
for(site in sites) {
  df <- divdfs[[site]]
  for(meth in meths) {
    dfkey <- paste0(site,".",meth)
    key <- paste(site,meth)
    df <- split_betas[[dfkey]]
    pv <- pvs[[meth]]
    pval <- mbs[[which(mbs$site==site & mbs$meth==meth), "pval"]]

    # LM modeling/plotting
    pb <- ggplot(data=df, aes(x=GWColl)) + geom_point(aes(y=beta), size=0.5)
    pb <- pb + geom_line(aes(y=pred), color="green")
    pb <- pb + geom_ribbon(aes(ymin=pred-2*SE, ymax=pred+2*SE), alpha=0.2, fill="green")
    pb <- pb + ylab(pv$ylab) + ylim(pv$ymin, pv$ymax)
    pb <- pb + xlab("Gestational Time (Weeks)")
    pb <- pb + annotate("text", x=pv$xan, y=pv$yan, size=ANSIZE, label = paste("p =", substr(as.character(pval),1,PVALLEN)))
    pb <- pb + ylab(NULL) + xlab(NULL)
    b_fits[[key]] <- pb
  }
}
library("gridExtra")
listorder <- paste(sites, rep(meths, each=4))
b_fits <- b_fits[listorder]
do.call(grid.arrange,c(b_fits, ncol=4, main=paste("Beta-diversity:", "Bray,", "JSD,", "WUF")))
```

# Make Publication Figures

Combine and label the generated plots to produce Figure 1:
```{r plot-main-text, warning=FALSE, message=FALSE}
ylab1 <- textGrob("Alpha-diversity\n(Shannon)", rot=90, gp=gpar(fontsize=8, col="black"))
ylab2 <- textGrob("Instability\n(w-Unifrac)", rot=90, gp=gpar(fontsize=8, col="black"))
ylab3 <- textGrob("Beta-diversity\n(w-Unifrac)", rot=90, gp=gpar(fontsize=8, col="black"))
xlab1 <- textGrob("          Gestational Weeks", gp=gpar(fontsize=8, col="black"))
title1 <- textGrob("      Vagina", gp=gpar(fontsize=8, col="black"))
title2 <- textGrob("       Stool", gp=gpar(fontsize=8, col="black"))
title3 <- textGrob("      Saliva", gp=gpar(fontsize=8, col="black"))
title4 <- textGrob("      Tooth/Gum", gp=gpar(fontsize=8, col="black"))
shanscale <- scale_y_continuous(limits=c(0,5), breaks=c(0,1.5,3,4.5))
ptheme <- theme(text = element_text(size=7),
                axis.title.x = element_text(size=8),
                axis.title.y = element_text(size=8),
                axis.text.x = element_text(size=7),
                axis.text.y = element_text(size=7),
                plot.title = element_text(size=8),
                plot.margin=unit(c(1,1,1,3),"mm"))
## MAIN FIGURE
Fig1 <- arrangeGrob(grob(), title1, title2, title3, title4,
      ylab1, ggs_alpha[[paste("Vaginal_Swab", "Shannon")]]+ylab(NULL)+xlab(NULL)+ggtitle(NULL)+shanscale + ptheme,
      ggs_alpha[[paste("Stool", "Shannon")]]+xlab(NULL)+ggtitle(NULL)+shanscale + ptheme,
      ggs_alpha[[paste("Saliva", "Shannon")]]+xlab(NULL)+ggtitle(NULL)+shanscale + ptheme, 
      ggs_alpha[[paste("Tooth_Gum", "Shannon")]]+xlab(NULL)+ggtitle(NULL)+shanscale + ptheme,
      ylab2, ggs_instability[[paste("Vaginal_Swab", "WUF")]]+xlab(NULL) + ptheme,
      ggs_instability[[paste("Stool", "WUF")]]+xlab(NULL) + ptheme,
      ggs_instability[[paste("Saliva", "WUF")]]+xlab(NULL) + ptheme, 
      ggs_instability[[paste("Tooth_Gum", "WUF")]]+xlab(NULL) + ptheme,
      ylab3, b_fits[[paste("Vaginal_Swab", "wuf")]]+ylab(NULL) + ptheme, 
      b_fits[[paste("Stool", "wuf")]]+xlab(NULL) + ptheme,
      b_fits[[paste("Saliva", "wuf")]]+xlab(NULL) + ptheme, 
      b_fits[[paste("Tooth_Gum", "wuf")]]+xlab(NULL) + ptheme,
      grob(), xlab1, xlab1, xlab1, xlab1,
      ncol=5, widths=unit.c(unit(2,"line"),rep(unit(1,"null"), 4)),
      nrow=5, heights=unit.c(unit(1,"line"), rep(unit(1,"null"), 3), unit(1,"line")))
Fig1
# ggsave(Fig1, file = "_PregPNAS_Fig1.pdf", width=17.5, height=9, units=c("cm"))
```

Combine and label the plots to make the Supplementary Figures:
```{r plot-supplement, warning=FALSE, message=FALSE}
ylab1 <- textGrob("Shannon", rot=90, gp=gpar(fontsize=12, col="black"))
ylab2 <- textGrob("Chao1", rot=90, gp=gpar(fontsize=12, col="black"))
ylab3 <- textGrob("Simpson", rot=90, gp=gpar(fontsize=12, col="black"))
xlab1 <- textGrob("Gestational Weeks", gp=gpar(fontsize=12, col="black"))
title1 <- textGrob("Vagina", gp=gpar(fontsize=12, col="black"))
title2 <- textGrob("Stool", gp=gpar(fontsize=12, col="black"))
title3 <- textGrob("Saliva", gp=gpar(fontsize=12, col="black"))
title4 <- textGrob("Tooth/Gum", gp=gpar(fontsize=12, col="black"))
shanscale <- scale_y_continuous(limits=c(0,5), breaks=c(0,1.5,3,4.5))
# SUPPLEMENTARY FIGURE
grid.arrange(grob(), title1, title2, title3, title4,
      ylab1, ggs_alpha[[paste("Vaginal_Swab", "Shannon")]]+ylab(NULL)+xlab(NULL)+ggtitle(NULL)+shanscale,
      ggs_alpha[[paste("Stool", "Shannon")]]+xlab(NULL)+ggtitle(NULL)+shanscale,
      ggs_alpha[[paste("Saliva", "Shannon")]]+xlab(NULL)+ggtitle(NULL)+shanscale, 
      ggs_alpha[[paste("Tooth_Gum", "Shannon")]]+xlab(NULL)+ggtitle(NULL)+shanscale,
      ylab2, ggs_alpha[[paste("Vaginal_Swab", "Chao1")]]+ylab(NULL)+xlab(NULL)+ggtitle(NULL),
      ggs_alpha[[paste("Stool", "Chao1")]]+xlab(NULL)+ggtitle(NULL),
      ggs_alpha[[paste("Saliva", "Chao1")]]+xlab(NULL)+ggtitle(NULL), 
      ggs_alpha[[paste("Tooth_Gum", "Chao1")]]+xlab(NULL)+ggtitle(NULL),
      ylab3, ggs_alpha[[paste("Vaginal_Swab", "Simpson")]]+ylab(NULL)+xlab(NULL)+ggtitle(NULL),
      ggs_alpha[[paste("Stool", "Simpson")]]+xlab(NULL)+ggtitle(NULL),
      ggs_alpha[[paste("Saliva", "Simpson")]]+xlab(NULL)+ggtitle(NULL), 
      ggs_alpha[[paste("Tooth_Gum", "Simpson")]]+xlab(NULL)+ggtitle(NULL),
      grob(), xlab1, xlab1, xlab1, xlab1,
      ncol=5, widths=unit.c(unit(1,"line"),rep(unit(1,"null"), 4)),
      nrow=5, heights=unit.c(unit(1,"line"), rep(unit(1,"null"), 3), unit(1,"line")))

ylab1 <- textGrob("Bray", rot=90, gp=gpar(fontsize=12, col="black"))
ylab2 <- textGrob("JSD", rot=90, gp=gpar(fontsize=12, col="black"))
ylab3 <- textGrob("w-Unifrac", rot=90, gp=gpar(fontsize=12, col="black"))
# SUPPLEMENTARY FIGURE
grid.arrange(grob(), title1, title2, title3, title4,
      ylab1, ggs_instability[[paste("Vaginal_Swab", "Bray")]]+ylab(NULL)+xlab(NULL)+ggtitle(NULL),
      ggs_instability[[paste("Stool", "Bray")]]+xlab(NULL)+ggtitle(NULL),
      ggs_instability[[paste("Saliva", "Bray")]]+xlab(NULL)+ggtitle(NULL), 
      ggs_instability[[paste("Tooth_Gum", "Bray")]]+xlab(NULL)+ggtitle(NULL),
      ylab2, ggs_instability[[paste("Vaginal_Swab", "JSD")]]+xlab(NULL),
      ggs_instability[[paste("Stool", "JSD")]]+xlab(NULL),
      ggs_instability[[paste("Saliva", "JSD")]]+xlab(NULL), 
      ggs_instability[[paste("Tooth_Gum", "JSD")]]+xlab(NULL),
      ylab3, ggs_instability[[paste("Vaginal_Swab", "WUF")]]+ylab(NULL)+xlab(NULL), 
      ggs_instability[[paste("Stool", "WUF")]]+xlab(NULL),
      ggs_instability[[paste("Saliva", "WUF")]]+xlab(NULL), 
      ggs_instability[[paste("Tooth_Gum", "WUF")]]+xlab(NULL),
      grob(), xlab1, xlab1, xlab1, xlab1,
      ncol=5, widths=unit.c(unit(1,"line"),rep(unit(1,"null"), 4)),
      nrow=5, heights=unit.c(unit(1,"line"), rep(unit(1,"null"), 3), unit(1,"line")))

ylab1 <- textGrob("Bray", rot=90, gp=gpar(fontsize=12, col="black"))
ylab2 <- textGrob("JSD", rot=90, gp=gpar(fontsize=12, col="black"))
ylab3 <- textGrob("w-Unifrac", rot=90, gp=gpar(fontsize=12, col="black"))
# SUPPLEMENTARY FIGURE
grid.arrange(grob(), title1, title2, title3, title4,
      ylab1, b_fits[[paste("Vaginal_Swab", "bray")]]+ylab(NULL)+xlab(NULL)+ggtitle(NULL),
      b_fits[[paste("Stool", "bray")]]+xlab(NULL)+ggtitle(NULL),
      b_fits[[paste("Saliva", "bray")]]+xlab(NULL)+ggtitle(NULL), 
      b_fits[[paste("Tooth_Gum", "bray")]]+xlab(NULL)+ggtitle(NULL),
      ylab2, b_fits[[paste("Vaginal_Swab", "jsd")]]+xlab(NULL),
      b_fits[[paste("Stool", "jsd")]]+xlab(NULL),
      b_fits[[paste("Saliva", "jsd")]]+xlab(NULL), 
      b_fits[[paste("Tooth_Gum", "jsd")]]+xlab(NULL),
      ylab3, b_fits[[paste("Vaginal_Swab", "wuf")]]+ylab(NULL), 
      b_fits[[paste("Stool", "wuf")]]+xlab(NULL),
      b_fits[[paste("Saliva", "wuf")]]+xlab(NULL), 
      b_fits[[paste("Tooth_Gum", "wuf")]]+xlab(NULL),
      grob(), xlab1, xlab1, xlab1, xlab1,
      ncol=5, widths=unit.c(unit(1,"line"),rep(unit(1,"null"), 4)),
      nrow=5, heights=unit.c(unit(1,"line"), rep(unit(1,"null"), 3), unit(1,"line")))
```

## Ordination of Communities During Pregnancy

Another way to look for systematic trends during pregnancy is to ordinate, and indicate time via color, allowing the eye to pick out (or not) trends.

Calculate ordination:
```{r ordinate, message=FALSE}
ords <- list()
for(site in sites) {
  ps <- PSPreg[[site]]
  ords[[site]] <- ordinate(ps, "NMDS", "bray")
}
``` 

Plot supplementary figure of ordinations for the different body sites:
```{r plot-ordinations}
require("RColorBrewer")
timepal <- colorRampPalette(brewer.pal(9,"RdBu"))(41)

contours <- list()
for(site in sites) {
  ps <- PSPreg[[site]]
  pord <- plot_ordination(ps, ords[[site]], type = "samples")
#  print(pord + geom_point(aes(color=GWColl), size = 3, alpha=0.5) + ggtitle(paste(site, "NMDS", "bray")))
  data <- pord$data
  data$Time <- c(NA)
  data$Time[data$GWColl %in% seq(14,21)] <- "Early"
  data$Time[data$GWColl %in% seq(28,35)] <- "Late"
  cont <- ggplot(data=data, aes(x=NMDS1, y=NMDS2))
  cont <- cont + geom_point(size=0.5) + geom_density2d(data=data[!is.na(data$Time),], aes(color=Time)) + xlim(-2,2) + ylim(-2,2)
  contours[[site]] <- cont + ggtitle(site)
}
# SUPPLEMENTARY FIGURE
do.call(grid.arrange, c(contours, ncol=2))
```

There is no trend evident in this plot, again supporting the idea that there is little systematic change in these communities over the course of pregnancy.
