---
title: "Course Activities"
author: "Marc A.T. Teunis"
date: "3/27/2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

```{r, include=FALSE}
exercise_files <- list.files(path = file.path(root, "exercises"), full.names = TRUE, recursive = TRUE)

presentation_files <- list.files(path = file.path(root, "presentations"), full.names = TRUE, recursive = TRUE)

##exercise_files
```

## Day 1

On day one we wil get introduced to Git, Github, Visualization examples in R RStudio,  and R basics and in that order. Why? 

 - Git is used to keep track of changes, collaborate and reproduce.
 - Github is used to share code (any code), data and other files and find additional R resources and packages.  

## Day 1 Visualizations

 - Are a great way to get inspired on the possibilities for using R
 - Are great for exploring new datasets
 - Are a great way to learn (new) ways of writing R code

## Day 1 RStudio

 - We will use RStudio Server as an Integrated Development Environment (IDE) for developing in R
 - RStudio is one of most widely used IDEs
 - RStudio integrates with the `Tidyverse`
 - RStudio enables easy creation of your own R package

## Day 1 Intro to R




## Day 1 files







