---
title: "Relative paths - {here}"
author: "Marc A.T. Teunis"
date: "4/1/2019"
output: html_document
---

## Project root

In most of the Rmd files used during the course, you will find this code:

```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

This will find the root of the project, even when the Rmd file is nested somewhere deep in you project. The Rmd file creates it's own environment, therefore it will not work if you use absolute paths in an Rmd file. The file will not render properly. 

## The `{here}` package generates a stable and more general solution to acces files through relative paths.

## **EXERCISE 1**

A) Go over the help for the `{here}` package
```{r}
library(here)
?here
```

B) Load the file "data/messy/TB_burden_countries_2016-12-08_messy.scv" into yor R session, using an aboslute path.

C) Now load the same datafile as in B), but now use the `root` object fromt the  



