---
title: "Conventions, Code and Data Management"
author: "Marc A.T. Teunis"
date: "`r Sys.Date()`"
output:
  ioslides_presentation: 
    highlight: pygments
    widescreen: yes
---

```{r setup1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.width = 5, 
                      fig.height = 3,
                      warning = FALSE,
                      message = FALSE,
                      error = FALSE)
```

```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

## Packages
```{r}
library(usethis)
library(devtools)
library(gapminder)
library(nycflights13)
library(tidyverse)
```

## Code Style

 * The slides in the course contain a number of style and syntax conventions.
 * Code styles and syntax can vary
 * Be consistent

## Capitals vs lower case

 * Everything in lower case
 * I use `snake_case`
 * Many people use `camelCase`

```{r}
gapminder %>% names
flights %>% names
```

## Printing a tbl_df
```{r}
table <- tibble("a_is_a_variable" = c(1:5), "b_is_another_variable" = rnorm(5))
table
```

## Variables

 * All lower case
 * Don't use capitals and special characters in file names

"`excel_file_of_important_dataset.xlsx`"

DON'T: "`data file + version 12_11_2017%.xlsx`"

## Packages referrals

Names of R packages are indicated as:

 * mostly: `{dplyr}`
 * sometimes: {dplyr}
 * not often: "dplyr"

## Explicit functions

When we want to be explicit about a certain function in a package:

`dplyr::filter()`

Refers to the `filter()` function in the package `{dplyr}`

## Loose code in the slides
Sometimes a few lines of codes are displayed for demo purpose:

`four_numbers <- c(1:4)`

`multiply <- 4*four_numbers`

`sum(multiply^2)`

## Code blocks

Code blocks in slides are indicated as below:
The result of the code is usually also displayed
```{r, fig.height=2, fig.width=3}
set.seed(1234)
code_that_does_something <- rnorm(1000)
hist(code_that_does_something, main = "rnorm")
```

## Hiding code 

Sometimes it makes sense to hide the code and only display the result. 
This can be done by setting the code chunk options `echo = FALSE`

```{r, echo=FALSE, fig.height=2.5, fig.width=3}
set.seed(1234)
code_that_does_something <- rnorm(1000)
hist(code_that_does_something, main = "rnorm")
```

## Other important chunk options

 * Like hiding the code, we can hide the result of that code and only display the code by setting `results = "hide"`
 * Setting `eval = FALSE` means the chunk is not included in the slide show at all
 * Setting `include = FALSE` means the chunk is run and the results are evaluated, but both the code and the results are not included in the slide-show/document 
 * Controlling settings for the complete document can be done in the `setup*` chunk that you will find in all Rmarkdown files from the course

```{r, setup2, eval = FALSE}
 knitr::opts_chunk$set(echo = TRUE,
                       message = FALSE,
                       warning = FALSE,
                       error = TRUE,
                       fig.width = 5, fig.height = 3)
```
 
 * There are many chunk options. Some we will see in later lessons

## Data and file management

 * Data for can be found in the "./data" folder of the course project
 * Sometime data needs to be downloaded from the internet. Usually the code is provided, but sometimes you will have to write a few lines of code to download data for e.g. an exercise or a case study. 
 * Many examples are with build-in data (e.g. `mtcars`, `gapminder`)

## Directions for recieving and gathering data

* **NEVER EVER** change a value of an original (RAW) dataset
* Add new variables to the dataset with recognizable names:
* Never change a file name of a raw dataset; this will cause tracibility of that file to break.

## Replacing old, creating new variables
```{r, eval=FALSE}
## replacing lifeExp in years for lifeExp in months
gapminder$lifeExp <- (gapminder$lifeExp*24)

## instead create a new variable in the dataframe
gapminder <- gapminder %>%
  mutate(lifeExp_months = lifeExp * 24)
```

_Mutate new varibles on the basis of old one, do not replace the old variable_

## Data Governance 

WIKIPEDIA: 12-05-2017:

 * Is a control that ensures that the data entry by an operations team member or by automated processes meets precise standards, 
 * The data governor uses data quality monitoring against production data to communicate errors in data for corrective action. 
 * Is used by organizations to exercise control over processes and methods used by their data stewards and data custodians in order to improve data quality.
 * Is a set of processes that ensures that important data assets are formally managed throughout the enterprise. 
 * Esures that data can be trusted and that people can be made accountable for any adverse event that happens because of low data quality. 
 * Is about putting people in charge of fixing and preventing issues with data so that the enterprise can become more efficient. 
 * Also describes an evolutionary process for a company, altering the company’s way of thinking and setting up the processes to handle information so that it may be utilized by the entire organization. 
 * It’s about using technology when necessary in many forms to help aid the process. When companies desire, or are required, to gain control of their data, they empower their people, set up processes and get help from technology to do it

## Programming in Data Govenance

 * Programming tools like R can aid governance of data
 * It helps debudding and finding errors, 
 * But can also introduce errors!!
 * Code is reproducible and sharable
 * Code helps communication about data and what happened to the data (in terms of gathering, cleaning, combining, filtering, visualizing)

## R packages

 * R packages are self-contained pieces of R code that allow easy sharing of R implementations
 * They follow a strict structure
 * They allow for versioning
 * The can be tested (unit tests)
 * They can be distributed (CRAN, Bioconductor, Github, Bitbucket, ...)
 * They can be made to fit multiple Operating Sytems
 
## There are many R packages

```{r, r_packages_graph, echo=FALSE}
source(here::here("/code/r_packages_graph.R"))
```
 
## Building an R package
During the course we will gradually build our own R package as a demonstration that it is actually not so difficult as it sounds.

## Initiate a package
Initiate a package skeleton with the following command:
```{r, eval=FALSE}  
library(usethis)
usethis::create_package(path = file.path("../courser"))
```
A new RStudio Project Session will open and will provide you with a skeleton of a new package called "courser". 

## Adding content to the package

As the course continues we will start filling this package skeleton with functions and data.

We will take three more steps to start with our package:

## 1. Package description
Look at the file named "DESCRIPTION". Provide some basic            information about yourself and what the package is intended for. Fill  in the fields:

  - `Title`; Provide a short description
  - `Authors@R`; provide your name and email-address (if you want)
  Leave the rest unchanged

## 2. Add a package dependency
Let's add our first dependency. The first dependecy that we will add is a package that we will use for data manipulation. It is called `{dplyr}` and we will use it very often. It makes sense to add this package as a dependency to our own package. The following code does that
```{r, eval=FALSE}
usethis::use_package(package = "dplyr")
```
 
## 3. Building a package
To build the package
run 
```{r, eval=FALSE}
devtools::document(pkg = file.path("../courser"))
```
Than run:
```{r, eval=FALSE}
system("R CMD INSTALL --preclean --no-multiarch --with-keep.source ../courser")
```

## **CONGRATULATIONS!!** 

You have juist build your own R package. We can now start adding functionality to it. We will do so as the course progresses.

 



